Early Detection of MCI that progresses to dementia - An Interdisciplinary Approach

### 2019-08-22 Source/Paper 2 - Systematic Review/Systematic Review.tex
updated Thesis

This SLR builds upon these questions and additionally presents the results of the other two additional research questions. Further, the key terms related to MCI and the other names by which it is known were included in the search string to ensure that relevant studies about machine learning and the diagnosis of a disease were also retrieved, even if not specifically mentioned in the paper's title or abstract.
\par 
To address the research questions, a search string was defined using the PICO approach, which decomposes the main research question into four parts: 
\begin{enumerate}
	\item Population - Studies that present research on mild cognitive impairment and dementia. Mild Cognitive Impairment and Dementia keywords were selected from the Systematized Nomenclature of Medicine–Clinical Terms (SNOMED-CT).
	\item Intervention - Intervention: ML or MS techniques. The ML keywords were selected from the branch “Machine Learning Approaches” of the “2012 ACM Computing Classification System”. The MS keywords were selected by A2.
	\item Comparison
	\item Outcome - Outcome: Prognosis on dementia and comorbidies. The prognosis keywords were provided
by A4.
\end{enumerate}

Using this process, the main research question was decomposed  into four research questions:
\begin{itemize}
	\item RQ1: Which NLP and ML techniques are being used in dementia research?
	\item RQ2: What features / data characteristics of text (variables, determinants and indicators) that are considered when applying the ML techniques (n-grams, PoS Tagging etc)?
	\item RQ3: What are the goals of the studies that employ NLP / ML techniques for prognosis of dementia?
	\item RQ4: Do the studies focus on time as factor?
\end{itemize}
\par 
The automated searches were performed in the Pubmed, Web of Science, Scopus and IEEE databases. Table 1 shows the search string used for the Pubmed automated search, but note that this search string was adapted to each of the other databases’ search context.

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ p{12cm} }
	\hline
	(dementia OR MCI OR Mild Cognitive Impairment OR Alzheimer's OR Mild Neurocognitive Disorder OR AD) AND TOPIC: (machine learning OR Data Mining OR Decision Support System OR NLP OR Natural 			Language Processing) AND TOPIC: (prognosis OR prognostic estimate OR predictor OR prediction OR model OR patterns OR diagnosis OR diagnostic OR forecasting OR projection OR Deep Language Model 		OR Deep Neural Network) AND TOPIC: (classification OR regression OR kernel OR support vector machines OR Gaussian Process OR Bayesian Network OR Factor Analysis OR Deep Learning OR Neural 			Networks OR Maximum Likelihood OR Principal Component Analysis OR Markov OR Linear Model OR Mixture Model OR Perceptron Algorithm OR Logical Learning OR relational learning OR Supervised 				Learning OR Unsupervised Learning OR clustering OR Decision Tree) AND TOPIC: (Language OR Cognitive OR Speech OR Conversation OR Connected Speech OR Picture Description OR Discourse Analysis OR 		Verbal Fluency)  \\ \hline
	Searched - 4th April 2019 - Generated 1257 Articles \\
	\hline
	\end{tabular}
	\end{center}
	\caption[Table caption text]{Example of Search Terms for Web of Science database}
	\label{table:name}
\end{table}

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ p{3cm} | p{3cm} | p{3cm} | p{3cm} }
	\hline
	Scopus & Web of Science & PubMed & IEEE Xplore \\ \hline
	1002 & 991 & 376 & 230 \\ 
	\hline
	\end{tabular}
	\end{center}
	\caption[Table caption text]{Number of articles found from each database - complete search terms appear in Appendix B}
	\label{table:name}
\end{table}

@@

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ p{6cm} | p{6cm} }
	\hline
	Inclusion Criteria & Exclusion Criteria \\ \hline
	Be a primary study in English; AND address research on dementia and comorbidities; AND address at least one ML or MS technique; AND address a prognosis related to dementia and comorbidities; AND use cognition or language decline as a factor for analysis. &
	Be a secondary or tertiary study; OR be written in another language other than English; OR do not address a research on dementia and comorbidities; OR do not address at least one ML or MS technique; OR do not address a prognosis related to dementia and comorbidities. \\
	\hline 
	\end{tabular}
	\caption[Table caption text]{Inclusion and Exclusion Criteria}
	\label{table:name}
	\end{center}
\end{table}

@@

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ p{6cm} | p{6cm} }
	\hline
	Variable & Definition \\ \hline
	Conditions Studied & For which dementia disorder is the study deriving a prognosis. \\ \hline
	Database used in the study & Name and origin of the data source used to derive the prognosis of the studied dementia. \\ \hline
	Dataset Categories & Classes in which the data units were divided into. \\ \hline
	Follow-up period & Period of time, which the data units were followed. \\ \hline
	Techniques used & Natural Language techniques AND/OR ML techniques that were used to build the diagnostic models. \\ \hline 
	Features generated & NLP generated features used in building the diagnostic models. \\ \hline
	Aim of the Study & The goal of the built diagnostic models. \\
	\hline 
	\end{tabular}
	\caption[Table caption text]{Quality Assessment Questionnaire}
	\label{table:name}
	\end{center}
\end{table}

@@

\section{Results}\label{results}
\subsection{Machine Learning methods}
Intro - Addresses Research Question 1

\subsubsection{Decision Trees}
A Decision Tree (DT) is a classification algorithm in which the learned knowledge is represented in a tree structure that can be translated to if-then rules. DT's learning process is recursive
and starts by testing each input variable as to how well each of them, alone, can classify the labeled examples. The best one is selected as a root node for the tree and its descendant nodes
are defined as the possible values (or relevant ratios) of the selected input variable. The training set is then classified between the descendant nodes according to the values of the selected input variable. This process is repeated recursively until no more splits in the tree are possible. Like SVMs, DTs do not depend on a pre-defined model and are mostly used
to find important interactions between variables. Being intuitive and easy to interpret, DTs have been used in prognostic studies as a tool for determining prognostic subgroups.
In this SLR, DTs were the second most frequently used ML technique, present in 6/37 selected studies and proposed in 7 models. It was employed for the same reason as SVM; except for one study that investigated the evolution of patients diagnosed with cognitive impairment no dementia (CIND) to AD.

\subsection{Features of Language}
Intro - Addresses Research Question 2

@@

\subsection{What are the goals of the studies that employ ML or MS techniques for prognosis of dementia and comorbidities?}

>>>>

This SLR builds upon these questions and additionally presents the results of the other two additional research questions. Further, the key terms related to MCI and the other names by which it is known were included in the search string to ensure that relevant studies about machine learning and the diagnosis of a disease were also retrieved, even if not specifically mentioned in the paper's title or abstract.
\par 
To address the research questions, a search string was defined using the PICO approach, which decomposes the main research question into four parts: 
\begin{enumerate}
	\item Population - Studies that present research on mild cognitive impairment and dementia. Mild Cognitive Impairment and Dementia keywords were selected from the Systematized Nomenclature of Medicine–Clinical Terms (SNOMED-CT).
	\item Intervention - Intervention: ML or Statistical Modelling techniques that focus on classification. The ML keywords were selected from the branch “Machine Learning Approaches” of the “2012 ACM Computing Classification System”. The MS keywords were selected by A2.
	\item Comparison
	\item Outcome - Outcome: Prognosis on dementia and comorbidies. The prognosis keywords were provided
by A4.
\end{enumerate}

Using this process, the main research question was decomposed  into four research questions:
\begin{itemize}
	\item RQ1: What features / data characteristics of text (variables, determinants and indicators) that are considered when applying the ML techniques (n-grams, PoS Tagging etc)?
	\item RQ2: Which NLP and ML or Statistical Learning techniques are being used in dementia research?
	\item RQ3: What are the goals of the studies that employ NLP / ML techniques for prognosis of dementia?
	\item RQ4: Do the studies focus on time as factor?
\end{itemize}
\par 
The automated searches were performed in the Pubmed, Web of Science, Scopus and IEEE databases. Table 1 shows the search string used for the Pubmed automated search, but note that this search string was adapted to each of the other databases’ search context.

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ | p{12cm} | }
	\hline
	(dementia OR MCI OR Mild Cognitive Impairment OR Alzheimer's OR Mild Neurocognitive Disorder OR AD) AND TOPIC: (machine learning OR Data Mining OR Decision Support System OR NLP OR Natural 			Language Processing) AND TOPIC: (prognosis OR prognostic estimate OR predictor OR prediction OR model OR patterns OR diagnosis OR diagnostic OR forecasting OR projection OR Deep Language Model 		OR Deep Neural Network) AND TOPIC: (classification OR regression OR kernel OR support vector machines OR Gaussian Process OR Bayesian Network OR Factor Analysis OR Deep Learning OR Neural 			Networks OR Maximum Likelihood OR Principal Component Analysis OR Markov OR Linear Model OR Mixture Model OR Perceptron Algorithm OR Logical Learning OR relational learning OR Supervised 				Learning OR Unsupervised Learning OR clustering OR Decision Tree) AND TOPIC: (Language OR Cognitive OR Speech OR Conversation OR Connected Speech OR Picture Description OR Discourse Analysis OR 		Verbal Fluency)  \\ \hline
	Searched - 4th April 2019 - Generated 1257 Articles \\
	\hline
	\end{tabular}
	\end{center}
	\caption[Table caption text]{Example of Search Terms for Web of Science database}
	\label{table:name}
\end{table}

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ | p{3cm} | p{3cm} | p{3cm} | p{3cm} | }
	\hline
	Scopus & Web of Science & PubMed & IEEE Xplore \\ \hline
	1002 & 991 & 376 & 230 \\ 
	\hline
	\end{tabular}
	\end{center}
	\caption[Table caption text]{Number of articles found from each database - complete search terms appear in Appendix B}
	\label{table:name}
\end{table}

@@

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ | p{6cm} | p{6cm} | }
	\hline
	Inclusion Criteria & Exclusion Criteria \\ \hline
	Be a primary study in English; AND address research on dementia and comorbidities; AND address at least one ML or MS technique; AND address a prognosis related to dementia and comorbidities; AND use cognition or language decline as a factor for analysis. &
	Be a secondary or tertiary study; OR be written in another language other than English; OR do not address a research on dementia and comorbidities; OR do not address at least one ML or MS technique; OR do not address a prognosis related to dementia and comorbidities. \\
	\hline 
	\end{tabular}
	\caption[Table caption text]{Inclusion and Exclusion Criteria}
	\label{table:name}
	\end{center}
\end{table}

@@

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ | p{6cm} | p{6cm} | }
	\hline
	Variable & Definition \\ \hline
	Conditions Studied & For which dementia disorder is the study deriving a prognosis. \\ \hline
	Database used in the study & Name and origin of the data source used to derive the prognosis of the studied dementia. \\ \hline
	Dataset Categories & Classes in which the data units were divided into. \\ \hline
	Follow-up period & Period of time, which the data units were followed. \\ \hline
	Techniques used & Natural Language techniques AND/OR ML techniques that were used to build the diagnostic models. \\ \hline 
	Features generated & NLP generated features used in building the diagnostic models. \\ \hline
	Aim of the Study & The goal of the built diagnostic models. \\
	\hline 
	\end{tabular}
	\caption[Table caption text]{Quality Assessment Questionnaire}
	\label{table:name}
	\end{center}
\end{table}

@@

\section{Results}\label{results}
\subsection{Features of Language}
One of the first steps in building an accurate diagnostic model using language is to analyse the effectiveness of the various language features that are used.

@@

\section{Machine Learning methods}
The next stage of building statistical or Machine Learning models is to take some features and use these feature to build models able to classify participants into various categories. There are various types of classification models that have been used to tackle this problem, this section looks which models have been used in the current literature.

\subsubsection{Decision Trees}
A Decision Tree (DT) is a classification algorithm in which the learned knowledge is represented in a tree structure that can be translated to if-then rules. DT's learning process is recursive and starts by testing each input variable as to how well each of them, alone, can classify the labeled examples. The best one is selected as a root node for the tree and its descendant nodes are defined as the possible values (or relevant ratios) of the selected input variable. The training set is then classified between the descendant nodes according to the values of the selected input variable. This process is repeated recursively until no more splits in the tree are possible. Like SVMs, DTs do not depend on a pre-defined model and are mostly used to find important interactions between variables. Being intuitive and easy to interpret, DTs have been used in prognostic studies as a tool for determining prognostic subgroups.
\par 
In this SLR, DTs were the second most frequently used ML technique, present in 6/37 selected studies and proposed in 7 models. It was employed for the same reason as SVM; except for one study that investigated the evolution of patients diagnosed with cognitive impairment no dementia (CIND) to AD.

\subsection{What are the goals of the studies that employ ML or Statistical Learning techniques for diagnosis of MCI or AD?}

\subsection{Do the studies focus on a one point in time or looking at cognitive deterioration over time?}

### 2019-08-22 Source/Thesis/Chapter1.tex
updated Thesis

\section{Introduction}
Alzheimer's Disease (AD) is a neurodegenerative disease in which, from a physiological perspective, the brain develops neurofibrillary tangles and neuritic plaques along with the deterioration and loss of cortical neurons and synapses. Whilst a definitive diagnosis of Dementia can only be produced at post-mortem, there are a number of clinical indicators from psychological perspective that can indicate dementia is present. From a clinical psychology perspective, those who have dementia demonstrate cognitive deficits such as problems with episodic and semantic memory, organizing and planning, difficulties with language, problems with executive function and visuospatial deficits \cite{McKhann2011}. In addition, these symptoms are often accompanied by emotional problems such as depression and behavioural difficulties. 

@@

AD and other forms of dementia affect a significant proportion of the geriatric population in the world today and is currently the sixth leading cause of death in the US and was named the leading killer of women in the UK. According to a recent report commissioned by the Alzheimer's Society in 2015, they estimate the prevalence of AD in the UK at approximately 815,000 people. This represents 1 in 14 of those aged 65 or over and 1 in 79 of the general population \cite{AlzheimersSociety2014}. From a financial perspective, they estimate an annual spend of £4.3 billion of which approximately £85 million is spent solely on diagnosis and that the total impact of AD (excluding the costs associated with early onset dementia) is £26.3 billion annually. Globally, this picture is a lot bleaker. Another report by Alzheimer's Disease International suggests that in 2015 there were 46 million people with a diagnosis of dementia globally and that number is expected to hit 131.5 million by 2050 \cite{Prince2015}. The report also states that the worldwide cost of AD in 2018 is estimated to be in the region of one trillon US dollars.

@@

Despite this growing problem, at present there are no drugs that improve the prognosis of those suffering with AD. All the drugs that are on the market are designed to manage symptoms. Whilst there are numerous investigational drugs in development for the treatment of AD, a larger than normal percentage (99.6\%) of these drugs fail in clinical trials (in contrast to anti-cancer drugs which have a 80\% failure rate) \cite{Cummings2014}. Researchers have proposed that a possible reason for the lack of success is that the drugs treatments are initiated too far along in the progression of the disease and thus much of the degeneration of the brain has already taken place \cite{Cummings2014}. It is therefore important to focus AD at it's earliest stages which some literature describes as 'Mild Cognitive Impairment (MCI) due to AD'.

@@

One of the challenges of this approach is differentiating natural cognitive decline due to aging with decline due to a form of cognitive impairment or dementia. Albert and his team have worked to define clinical criteria which professionals can use to diagnose MCI due to AD and differentiate this from age-associated memory impairment and age-associated cognitive decline. They note that the diagnosis of MCI requires evidence of intra-individual change and optimally requires evaluation at two or more points \cite{Albert2011}. The criteria for MCI is detailed below \cite{Albert2011}.

@@

@@

@@

@@

This thesis' original contributions are an increased understanding of how language deteriorates over time in those with MCI and early AD, the exploration via a delphi method of ideas for the collection of language features via smartphone or other internet of things devices and the development and execution of a pilot study to test the feasibility of collecting high quality language samples via in-home technology. I also contribute to the research into which language features are important in the diagnosis of MCI and early AD over time. Finally, I explore the potential of one class classification as a new approach to classifying those with MCI and Early AD vs controls. The potential impact of this research in this area is immense. Research has shown that early diagnosis of people with AD or MCI improves sufferers quality of life and can, in some cases, slow the progress of the disease. Early diagnosis can increase the number of research opportunities for understanding the early stages of dementia and how the disease progresses so that more research can be conducted which may, in the future, lead to new treatments and other interventions.

>>>>

\section{Background}
Alzheimer's Disease (AD) and other forms of dementia affect a significant proportion of the geriatric population in the world today and is currently the sixth leading cause of death in the US and was named the leading killer of women in the UK. According to a recent report commissioned by the Alzheimer's Society in 2015, they estimate the prevalence of AD in the UK at approximately 815,000 people. This represents 1 in 14 of those aged 65 or over and 1 in 79 of the general population \cite{AlzheimersSociety2014}. From a financial perspective, they estimate an annual spend of £4.3 billion of which approximately £85 million is spent solely on diagnosis and that the total impact of AD (excluding the costs associated with early onset dementia) is £26.3 billion annually. Globally, this picture is a lot bleaker. Another report by Alzheimer's Disease International suggests that in 2015 there were 46 million people with a diagnosis of dementia globally and that number is expected to hit 131.5 million by 2050 \cite{Prince2015}. The report also states that the worldwide cost of AD in 2018 is estimated to be in the region of one trillon US dollars.

@@

AD is a neurodegenerative disease in which a definitive diagnosis can only be produced at post-mortem. However, there are a number of psychological and physiological indicators that can indicate that dementia is present. From a physiological perspective, researchers have identified two proteins called beta-amyloid and tau. In a typical case, tau accumulates and eventually forms tangles inside neurons and beta-amyloid clumps into plaques which slowly builds up between neurons. At a certain point, the levels of beta-amyloid rise and trigger a more rapid spread of tau throughout the brain. Eventually, due to this and other changes, neurons lose their ability to communicate and the brain starts to shrink. This leads to some of the more psychological symptoms, those who have dementia demonstrate cognitive deficits such as problems with episodic and semantic memory, organizing and planning, difficulties with language, problems with executive function and visuospatial deficits \cite{McKhann2011}. In addition, these symptoms are often accompanied by emotional problems such as depression and behavioural difficulties. As more neurons die throughout the brain, a person with Alzheimer's gradually loses the ability to think, remember, make decisions and function independently.

@@

Despite this growing problem and an increasing understanding about how AD affects the brain there are no medications that improve the prognosis of those with AD. All the medications that are currently on the market are designed to manage symptoms. Whilst there are numerous investigational drugs in development for the treatment of AD, a larger than normal percentage (99.6\%) of these drugs fail in clinical trials (in contrast to anti-cancer drugs which have a 80\% failure rate) \cite{Cummings2014}. Researchers have proposed that a possible reason for the lack of success is that the drugs treatments are initiated too far along in the progression of the disease and thus much of the degeneration of the brain has already taken place \cite{Cummings2014}. Research has started to be more focused on AD at it's earliest stages which some literature describes as 'Mild Cognitive Impairment (MCI) due to AD'.

@@

One of the challenges of this approach is differentiating natural cognitive decline due to aging with decline due to a form of cognitive impairment or dementia. This challenge is often complicated further due to the large variation in the cognitive abilities and educational background of individuals. Albert and his team have worked to define clinical criteria which professionals can use to diagnose MCI due to AD and differentiate this from age-associated memory impairment and age-associated cognitive decline.  One of the most important observations from this piece of work is that a diagnosis of MCI requires evidence of intra-individual change and optimally requires evaluation at two or more points \cite{Albert2011}, and this is essentially to place more importance on the trajectory of a person's cognitive abilities rather than a person's cognitive ability in general. The criteria for MCI is detailed below \cite{Albert2011}.

@@

Research has shown that early diagnosis of people with AD or MCI improves sufferers quality of life and can, in some cases, slow the progress of the disease. Early diagnosis can increase the number of research opportunities for understanding the early stages of dementia and how the disease progresses so that more research can be conducted which may, in the future, lead to new treatments and other interventions.

@@

\section{Statement of the Problem}

@@

\section{Research Questions}

@@

This thesis' original contributions are an increased understanding of how language deteriorates over time in those with MCI and early AD, the exploration via a delphi method of ideas for the collection of language features via smartphone or other internet of things devices and the development and execution of a pilot study to test the feasibility of collecting high quality language samples via in-home technology. I also contribute to the research into which language features are important in the diagnosis of MCI and early AD over time. Finally, I explore the potential of one class classification as a new approach to classifying those with MCI and Early AD vs controls. The potential impact of this research in this area is immense. 
\section{Structure of thesis}
